---
title: "Wavelet analysis on 14 high-burden countries and Vietnam"
toc: true
---

## Data processing

### Load data

```{r message=FALSE, warning=FALSE}
#| code-fold: true
#| code-summary: "Load measles epicurve data for SEA countries"
suppressMessages(library(tidyverse))
library(janitor)
library(plotly)
library(patchwork)
library(countrycode)

world_epi_curve <- readxl::read_excel(
  "data/404-table-web-epi-curve-data.xlsx",
  sheet = "WEB")

# --- preprocess ----- 
world_epi_curve <- world_epi_curve %>% 
  clean_names() %>% 
  mutate(
    date = ym(str_c(year, month)),
    measles_clinical = as.integer(measles_clinical)
  ) 

filtered_countries <- c(
  "Viet Nam", "India", "Nigeria", "Indonesia", "Ethiopia", "China", "Philippines", "Uganda", "DR Congo", "Pakistan", "Angola", "Madagascar","Ukraine", "Malawi", "Somalia")

# countries that introduce mcv2 to their routine vaccination before 2017
early_mcv2 <- c("India", "Indonesia", "China", "Philippines", "Pakistan", "Angola", "Ukraine", "Malawi", "Vietnam")
# countries that introduce mcv2 to their routine vaccination from 2017 - 2020
late_mcv2 <- c("Nigeria", "Ethiopia", "Madagascar")
# countries that introduce mcv2 to their routine vaccination as of 2020
no_mcv2 <- c("DR Congo", "Uganda","Somalia")

filtered_iso <- countrycode(filtered_countries, origin = "country.name", destination = "iso3c")

# get epi curve of high burden countries
highburden_epi_curve <- world_epi_curve %>% 
  clean_names() %>% 
  mutate(
    date = ym(str_c(year, month)),
    measles_suspect = as.integer(measles_suspect),
    measles_clinical = as.integer(measles_clinical),
    measles_epi_linked = as.integer(measles_epi_linked),
    measles_lab_confirmed = as.integer(measles_lab_confirmed),
    # NOTE: recompute measles total since some epi-linked cases are negative (e.g. record for Indonesia on 2016-10-01)
    # definition: t sum of clinically-compatible, epidemiologically linked and laboratory-confirmed cases
    measles_total = abs(measles_clinical) + abs(measles_epi_linked) + abs(measles_lab_confirmed), 
    # rename Congo for shorter name
    country = if_else(country == "Democratic Republic of the Congo", "DR Congo", country)
  ) %>% 
  filter(
    iso3 %in% filtered_iso
  ) 
```

```{r}
#| code-fold: true
#| code-summary: "helper functions"
to_iso <- \(.){countrycode(., origin = "country.name", destination = "iso3c")}
to_region <- \(.){countrycode(., origin = "country.name", destination = "region")}
```

### Clean data

-   The column used for incidence data is measles total (sum of clinical compatible cases, epi-linked cases, laboratory confirmed cases)

-   Make sure the the time series is even-spaced

-   Check % of missing data from each of the 11 SEA countries

```{r}
#| code-fold: true
#| code-summary: "Generate evenly spaced time series for all countries"
date_range <- range(highburden_epi_curve$date)
highburden_epi_curve <- highburden_epi_curve %>% 
  group_by(country) %>%
  rename(
    incidence = measles_total
  ) %>% 
  complete(
    date = seq(date_range[1], date_range[2], by = "month"),
    fill = list(
      incidence = 0
    )
  ) %>% 
  ungroup()
```

View the percentage of data points where `incidence==0`

```{r}
#| code-fold: true
#| code-summary: "Get country with most incidence report being 0"
percent_zero <- highburden_epi_curve %>% 
  group_by(country) %>% 
  summarize(
    percent_zero = sum(incidence == 0)/n()
  ) %>% 
  arrange(
    desc(percent_zero)
  ) 

percent_zero

# filtered out countries where percent of data points where incidence = 0 > 50%
high_pct_zero <- percent_zero %>% filter(percent_zero > 0.4) %>% pull(country)
```

### Plot incidence

```{r}
#| code-fold: true
#| code-summary: "Plor raw incidence"
highburden_epi_plot <- highburden_epi_curve %>% 
  filter(!(country %in% high_pct_zero)) %>% 
  ggplot() +
  geom_line(
    aes(
      x = date,
      y = incidence,
      color = country
    )
  ) 

ggplotly(highburden_epi_plot)
```

### Preprocess data

Follow these steps

-   Log transform

-   Normalized

-   Remove latest observation of each country (most likely under-reported)

```{r}
#| code-fold: true
#| code-summary: "Preprocess data"
filtered_epicurve <- highburden_epi_curve %>% 
  filter(!(country %in% high_pct_zero)) 

log_transformed <- filtered_epicurve %>% 
  group_by(country) %>% 
  mutate(
    # log transform
    incidence = log10(incidence + 1),
    # normalized
    incidence = (incidence - mean(incidence))/sd(incidence)
  ) %>% 
  filter(
    date != max(date)
  ) %>% 
  ungroup()
```

```{r}
#| code-fold: true
#| code-summary: "Plot preprocessed data"
log_epicurve_plot <- log_transformed %>% 
  ggplot() +
  geom_line(
    aes(
      x = date,
      y = incidence,
      color = country
    )
  ) 

ggplotly(log_epicurve_plot)
```

## Wavelet analysis

```{r message=FALSE, warning=FALSE}
library(WaveletComp)
```

Follow these steps of wavelet analysis

-   Inspect the **power spectrum across time-period domain**: identify the period that is consistently high over time, which indicates recurring cycles in the time series.

-   Inspect the **global power** (time-averaged power, analogue of the Fourier transform): identify the overall dominant periodicities.

-   Inspect the **period-specific reconstructed component**: extract period-specific oscillations to assess their contributions to the time series.

-   Inspect the **period-specific phase**: visualize the timing of cycles of different time series.

***General setup***

-   1 time unit is 1 year. Since the data is collected every month, `dt = 1/12`

-   The analyzed range of period is `[1/6, 8]` (i.e., 2 months to 8 years)

### Perform analysis on all data

```{r message = FALSE}
#| code-fold: true
#| code-summary: "Perform wavelet data analysis on high burden countries"
countries_wavelets <- log_transformed %>% 
  group_by(country) %>% 
  nest() %>% 
  mutate(
    wavelet = map(data, \(dat){
      capture.output(
        out <- dat %>% 
        analyze.wavelet(
          my.series = "incidence",
          dt = 1/12, # 1 time unit = 1 year 
          lowerPeriod = 1/6, # 2 months
          upperPeriod = 8, #  years
          verbose = FALSE
        )
      ) %>% invisible()
        
      out
    })
  ) %>% 
  select(country, wavelet) %>%
  deframe()
```

### Power spectrum

These plots suggest that high burden countries tend to experience annual or 5-year period

```{r}
#| code-fold: true
#| code-summary: "Function to plot power spectrum of all countries"

plot_power <- function(wavelets){
  # return heatmap of power and also global power
  imap(wavelets, \(wavelet, country){
    wt.image(
        wavelet,
        main = paste0(country, " (", to_region(country) ,")", " power spectrum"),
        legend.params = list(lab = "Power"),
        periodlab = "Period(Year)",
        show.date = TRUE
      )
    
    wt.avg(
        wavelet,
        main = paste0(country," (", to_region(country) ,")", " global power")
      )
  })
}
```

```{r}
plot_power(countries_wavelets)
```

### Prominent period

```{r}
#| code-fold: true
#| code-summary: "Prominent measles cycle of each country"
map2_dfr(names(countries_wavelets), countries_wavelets, \(name, out){
  data.frame(
    period = out$Period,
    power = out$Power.avg,
    p_val = out$Power.avg.pval,
    country = name
  ) %>% 
  filter(p_val <= 0.05) %>% 
  arrange(desc(power)) %>% 
  head(n=1)
}) %>% 
mutate(
  # label period
  rounded_period = round(period),
  region = to_region(country)
) %>% 
arrange(rounded_period, desc(power)) %>% 
select(rounded_period, country, region, power)
```

### Reconstructed component

```{r}
#| code-fold: true
#| code-summary: "Function to plot period specific component"
reconstruct_ts <- function(wavelets, period = 5){
  df <- map2_dfr(wavelets, names(wavelets), \(wavelet, country){
    rec_period <- reconstruct(
      wavelet,
      sel.period = period,
      plot.rec = FALSE,
      verbose = FALSE
    )
    
    rec_period$series %>%
      mutate(country = country, .before = 1)
  })
  
  plot <- df %>% 
    ggplot() +
      geom_line(
        aes(x = date, y = incidence.r, color = country)
      ) +
      labs(
        y = paste0(round(mean(period), digits = 2), "-year period component"),
        x = "Date"
      )
  
  # return data.frame and plot
  list(
    df = df,
    plot = plot
  )
}
```

::: panel-tabset
## 5-year

Take the average of the components of period ranging from 4-5.

Clear oscillations can be observed across all the countries, except Nigeria

```{r}
ggplotly(
  reconstruct_ts(countries_wavelets, period = c(4,5))$plot
)
```

## 2-year

Roughly biennial component. Take the average of the components of period ranging from 1.9-2.5.

Relatively consistent biennual cycle can be observed for: Viet Nam, Nigeria, Pakistan, DR Congo

```{r}
ggplotly(
  reconstruct_ts(countries_wavelets, period = c(1.9, 2.5))$plot
)
```

## 1-year

The plot suggests that:

-   Pakistan, Nigeria exhibits consistent annual cycle overtime

-   DR Congo exhibits relatively consistent annual cycle since 2017

-   Ethiopia, China, India, Somalia, Uganda experiences consistent annual measles cycle at the start but diminish over time

```{r}
ggplotly(
  reconstruct_ts(countries_wavelets, period = c(0.9,1.1))$plot
)
```
:::

### Phase

```{r}
#| code-fold: true
#| code-summary: "Function to plot period-specific phase"
compute_phase <- function(wavelets, period = 5){
  df <- map2_dfr(wavelets, names(wavelets), \(wavelet, country){
    
    # if only one period is given -> select the closest period
    # else select all period within the range
    row_period <- if(length(period) == 1) which.min(abs(wavelet$Period - period)) else
      which( (wavelet$Period >= min(period)) & (wavelet$Period <= max(period)) )
    
    data.frame(
      date = wavelet$series$date,
      # compute mean phase across all selected periods
      phase = if(length(row_period) > 1) colMeans(wavelet$Phase[row_period,]) else wavelet$Phase[row_period,]
    ) %>% 
      mutate(
        country = country,
        .before = 1
      )
  })
  
  plot <- df %>% 
    ggplot() +
      geom_line(
        aes(x = date, y = phase, color = country)
      ) +
      labs(
        y = "Phase",
        x = "Date"
      )
  
  # return data.frame and plot
  list(
    df = df,
    plot = plot
  )
}
```

```{r include=FALSE}
#| code-fold: true
#| code-summary: "Function to plot period-specific phase difference"

# Manually compute phase difference
# There will be an alternative function to make use package (WaveletComp) functions later on
compute_phase_diff_alt <- function(wavelets, period = 5, ref_country = "Viet Nam", 
                               constraint = TRUE, diff_to_time = TRUE){
  # compute phase of each of these countries
  phases <- compute_phase(wavelets, period = period)$df
  
  ref_country_phase <- phases %>% 
    filter(
      country == ref_country
    ) %>% 
    select(-country)
  
  # get vector of remaining country
  remaining_countries <- phases %>% filter(country != ref_country) %>% pull(country) %>% unique()
  
  df <- map_dfr(remaining_countries, \(curr_country){
    
    data.frame(
      date = ref_country_phase$date,
      phase_diff = ref_country_phase$phase - phases %>% 
            filter(
              country == curr_country
            ) %>% pull(phase)
    ) %>% 
    mutate(
      # constraint the phase difference within +-pi
      phase_diff = if(constraint) ((phase_diff + 3*pi) %% (2*pi)) - pi else phase_diff,
      phase_diff = if(diff_to_time) phase_diff*mean(period)/(2*pi) else phase_diff,
      label = paste0(ref_country, "-", curr_country)
    )
  })
  
  plot <- df %>% 
    ggplot() +
      geom_line(
        aes(x = date, y = phase_diff, color = label)
      ) +
      geom_hline(
        aes(yintercept = 0),
        color = "black", linetype = "dashed"
      ) + 
    labs(
      title = paste0("Phase difference between ", ref_country, " and other countries"),
      xlab = "Date",
      y = paste0("Phase difference (in ", if(diff_to_time) "time unit" else "radians", ")")
    )
    
  list(
    df = df, plot = plot
  )
}
```

::: panel-tabset
## 5-year

Major epidemics cycle (roughly 5 years)

```{r}
countries_phase <- compute_phase(countries_wavelets, period = 5)

ggplotly(countries_phase$plot)
```

## 1-year

Roughly annual cycle

```{r}
countries_phase <- compute_phase(countries_wavelets, period = 1)
ggplotly(countries_phase$plot)
```
:::

### Periodicity over time

Plot power of period specific oscillations over time. Different linetypes represent the time period when MCV2 was introduced into the routine vaccination program of the corresponding country.

```{r}
#| code-fold: true
#| code-summary: "Function to plot period-specific power over time"
plot_power_time <- function(wavelets, period = 5){
  df <- map2_dfr(names(wavelets), wavelets, \(country, wavelet){
    # if only one period is given -> select the closest period
    # else select all period within the range
    row_period <- if(length(period) == 1) which.min(abs(wavelet$Period - period)) else
      which( (wavelet$Period >= min(period)) & (wavelet$Period <= max(period)) )
    
    data.frame(
      date = wavelet$series$date,
      country = country,
      # compute mean power
      power = if(length(row_period) > 1) colMeans(wavelet$Power[row_period,]) else wavelet$Power[row_period,],
      mcv2_time = case_when(
        to_iso(country) %in% to_iso(early_mcv2) ~ "Before 2017",
        to_iso(country) %in% to_iso(late_mcv2) ~ "2017-2019",
        to_iso(country) %in% to_iso(no_mcv2) ~ "Unintroduced by 2020",
        .default = "Unknown"
      )
    )
  })
  
  plot <- df %>% 
    ggplot() +
    geom_line(
      aes(x = date, y = power, color = country, linetype = mcv2_time)
    ) + 
    labs(
      title = paste0("Power of ", mean(period), "-year period"),
      x = "Date",
      y = "Powers",
      linetype = "Year introduce MCV2"
    )
  
  list(
    df = df,
    plot = plot
  )
}
```

::: panel-tabset

## 5-year
Take the average power of period ranging from 4-5
```{r}
ggplotly(plot_power_time(countries_wavelets, period = c(4,5))$plot)
```


## 2-year
Take the average power of period ranging from 1.5-2.5
```{r}
ggplotly(plot_power_time(countries_wavelets, period = c(1.5,2.5))$plot)
```

## 1-year
Take the average power of period ranging from 0.8-1.2
```{r}
ggplotly(plot_power_time(countries_wavelets, period = c(0.8,1.2))$plot)
```

:::
